---
title: "Real_state"
author: "Ignacio Almodóvar, Luis Rodríguez, Javier Muñoz"
date: "12/4/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

First of all, we are going to read the Boston house prices dataset file and show the summary and the structure:

```{r}
dataset <- read.csv("Dataset.csv")

summary(dataset)
str(dataset)

```

Next step is to convert the type of the categorical variable _CHAS_ into factor and create a few categorical variables from suitable quantitative ones:

- _RAD.CAT_: low, medium and high
- _RM.CAT_: low, medium and high
- _INDUS.CAT_: very low, low, high and very high
- _LSTAT.CAT_: very low, low, high and very high
- _ZN.CAT_: very low, low, high and very high

```{r}

dataset$CHAS <- as.factor(dataset$CHAS)

factor_rad_values <- sort(unique(dataset$RAD))
delimiter_rad_index <- length(factor_rad_values)/3
lower_delimiter_rad_value <- factor_rad_values[round(delimiter_rad_index)]
upper_delimiter_rad_value <- factor_rad_values[round(2*delimiter_rad_index)]

dataset$RAD.CAT <- cut(dataset$RAD, 
    breaks = c(-Inf, lower_delimiter_rad_value, upper_delimiter_rad_value, Inf), 
    labels = c('low', 'medium', 'high'),
    right = FALSE)

dataset$RM.CAT <- cut(dataset$RM, 
    breaks = c(-Inf, 4, 6, Inf), 
    labels = c('low', 'medium', 'high'),
    right = FALSE)

factor_indus_values <- sort(unique(dataset$INDUS))
delimiter_indus_index <- length(factor_indus_values)/4
lower_delimiter_indus_value <- factor_indus_values[round(delimiter_indus_index)]
mid_delimiter_indus_value <- factor_indus_values[round(2*delimiter_indus_index)]
upper_delimiter_indus_value <- factor_indus_values[round(3*delimiter_indus_index)]

dataset$INDUS.CAT <- cut(dataset$INDUS, 
    breaks = c(-Inf, lower_delimiter_indus_value, mid_delimiter_indus_value, upper_delimiter_indus_value, Inf), 
    labels = c('very low', 'low', 'high', 'very high'),
    right = FALSE)

dataset$LSTAT.CAT <- cut(dataset$LSTAT, 
    breaks = c(-Inf, 25, 50, 75, Inf), 
    labels = c('very low', 'low', 'high', 'very high'),
    right = FALSE)

dataset$ZN.CAT <- cut(dataset$ZN, 
    breaks = c(-Inf, 25, 50, 75, Inf), 
    labels = c('very low', 'low', 'high', 'very high'),
    right = FALSE)

str(dataset)

```


