---
title: "Real_state"
author: "Ignacio Almodóvar, Luis Rodríguez, Javier Muñoz"
date: "12/4/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

First of all, we are going to read the Boston house prices dataset file and show the summary and the structure:

```{r}
library(ggplot2)
library(dplyr)
library(MASS)
dataset <- read.csv("Dataset.csv")
```

Variables explanation

Attribute Information:

- CRIM: This is a continuous variable that measures the crime per capital rate by town.
- ZN: Continuous variable that indicates the proportion of residential land zoned for lots over 25,000 sq.ft. There are not many different values in this category. Could be also considered a categorical variable.
- INDUS: Proportion of non-retail business acres per town. Continuous variable.
- CHAS: It is a binary variable referred to Charles River (= 1 if tract bounds river; 0 otherwise).
- NOX: Continuous variable that measures nitric oxides concentration (parts per 10 million).
- RM: Measures the average number of rooms per dwelling. Continuous variable.
- AGE: Continuous variable that measures the proportion of owner-occupied units built prior to 1940.
- DIS: Continuous variable that indicates the weighted distances to five Boston employment centers.
- RAD: Index of accessibility to radial highways. Larger index denotes better accessibility.
- TAX: Continuous variable full-value property-tax rate per $10,000.
- PTRATIO: Continuous variable that measures the proportion of pupil-teacher by town.
- B: Is a continuous variable based on the formula 1000(Bk - 0.63)^2, where BK is the proportion of black people.
- LSTAT: Continuous variable that measures the proportion of adults without, some high school education and proportion of male workers classified as laborers.
- MEDV: Continuous variable with the median value price of owner-occupied homes in $1000's


Next step is to convert the type of the categorical variable _CHAS_ into factor and create a few categorical variables from suitable quantitative ones:

- _RAD.CAT_: low, medium and high
- _RM.CAT_: low, medium and high
- _INDUS.CAT_: very low, low, high and very high
- _LSTAT.CAT_: very low, low, high and very high
- _ZN.CAT_: very low, low, high and very high

```{r}

dataset$CHAS <- as.factor(dataset$CHAS)

factor_rad_values <- sort(unique(dataset$RAD))
delimiter_rad_index <- length(factor_rad_values)/3
lower_delimiter_rad_value <- factor_rad_values[round(delimiter_rad_index)]
upper_delimiter_rad_value <- factor_rad_values[round(2*delimiter_rad_index)]

dataset$RAD.CAT <- cut(dataset$RAD, 
    breaks = c(-Inf, lower_delimiter_rad_value, upper_delimiter_rad_value, Inf), 
    labels = c('low', 'medium', 'high'),
    right = FALSE)

dataset$RM.CAT <- cut(dataset$RM, 
    breaks = c(-Inf, 4, 6, Inf), 
    labels = c('low', 'medium', 'high'),
    right = FALSE)

factor_indus_values <- sort(unique(dataset$INDUS))
delimiter_indus_index <- length(factor_indus_values)/4
lower_delimiter_indus_value <- factor_indus_values[round(delimiter_indus_index)]
mid_delimiter_indus_value <- factor_indus_values[round(2*delimiter_indus_index)]
upper_delimiter_indus_value <- factor_indus_values[round(3*delimiter_indus_index)]

dataset$INDUS.CAT <- cut(dataset$INDUS, 
    breaks = c(-Inf, lower_delimiter_indus_value, mid_delimiter_indus_value, upper_delimiter_indus_value, Inf), 
    labels = c('very low', 'low', 'high', 'very high'),
    right = FALSE)

dataset$LSTAT.CAT <- cut(dataset$LSTAT, 
    breaks = c(-Inf, 25, 50, 75, Inf), 
    labels = c('very low', 'low', 'high', 'very high'),
    right = FALSE)

dataset$ZN.CAT <- cut(dataset$ZN, 
    breaks = c(-Inf, 25, 50, 75, Inf), 
    labels = c('very low', 'low', 'high', 'very high'),
    right = FALSE)

str(dataset)

```



I'm going to try to find differences in shapes between the different levels of a predictor and the variable that we want to predict, the average price.



```{r}
pairs(data=dataset,~CRIM+ZN+INDUS+NOX+RM+AGE+DIS+RAD+TAX+PTRATIO+B+LSTAT+MEDV,pch=19,col=dataset$CHAS)
```

Within this plot we can see that there is a linear relationship between (NOX,DIS) and (LSTAT,MEDV)

Now, lets try to find groups through scatter plots.
```{r}
ggplot(data=dataset,aes(x=NOX,y=DIS))+geom_point(aes(colour=ZN.CAT))

ggplot(data=dataset,aes(y=MEDV,x=LSTAT))+geom_point(aes(colour=LSTAT.CAT))
```

Density curves

```{r}
#Given river
dataset %>% ggplot(aes(x=MEDV)) + geom_density(aes(color=CHAS))

#Given rooms
dataset %>% ggplot(aes(x=MEDV)) + geom_density(aes(color=RM.CAT,fill=RM.CAT))

dataset %>% ggplot(aes(x=MEDV)) + geom_density(aes(color=LSTAT.CAT))

dataset %>% ggplot(aes(x=(NOX))) + geom_density(aes(color=ZN.CAT))
```

PCP plot

```{r}
color_rad_cat=c("blue","green","orange2")[dataset$RAD.CAT]
parcoord(dataset[,c(1,5,7:8,11,12,14)],col=color_rad_cat,var.label=TRUE)
```
We can see that for high RAD.CAT the variables more affected are CRIM, AGE and DIS

ANDREWS curve

```{r}
library(andrews)
par(mfrow=c(1,1))
andrews(df=as.data.frame(cbind(dataset[,c(1,5,7:8,11,12,14)],dataset$CHAS)),clr=8,ymax=4)
```

###############################3

Checking for NA

```{r}
missingValues=function(data){
  count=0
  a=cbind(lapply(lapply(data, is.na), sum))
  for(i in 1:ncol(data)){
    if(a[i]!=0){
      cat(as.integer(a[i]), "missing values in column ", i,"\n" )
      count=count+1
    }
  }  
    if(count==0){
      cat("There are No missing values in this dataset")
    }
}

missingValues(dataset)
```

Impute NA

```{r}
#With the mean
clean_data_mean=dataset
index=which(is.na(dataset$RM))
clean_data_mean$RM[index]=mean(clean_data_mean$RM,na.rm = TRUE)
clean_data_mean$RM[index]

#Using mice library
library(mice)
set.seed(1234)
methods(mice)
imputation=mice(dataset,m=5,method="pmm",maxit = 20)

#Check each prediction
imputation$imp$RM

#We want to check something close to the mean value 6.288 so we are going to take the first prediction, because it has an error lower than 0.5 on each boundary
summary(dataset$RM)

for(i in 1:5){
    print(imputation$imp$RM[i]-clean_data_mean$RM[index])
}

clean_data=complete(imputation,1)
```


