---
title: "Real_state"
author: "Ignacio Almodóvar, Luis Rodríguez, Javier Muñoz"
date: "12/4/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

First of all, we are going to read the Boston house prices dataset file and show the summary and the structure:

```{r}
library(ggplot2)
library(dplyr)
library(MASS)
dataset <- read.csv("Dataset.csv")

summary(dataset)
str(dataset)

```

Next step is to convert the type of the categorical variable _CHAS_ into factor and create a few categorical variables from suitable quantitative ones:

- _RAD.CAT_: low, medium and high
- _RM.CAT_: low, medium and high
- _INDUS.CAT_: very low, low, high and very high
- _LSTAT.CAT_: very low, low, high and very high
- _ZN.CAT_: very low, low, high and very high

```{r}

dataset$CHAS <- as.factor(dataset$CHAS)

factor_rad_values <- sort(unique(dataset$RAD))
delimiter_rad_index <- length(factor_rad_values)/3
lower_delimiter_rad_value <- factor_rad_values[round(delimiter_rad_index)]
upper_delimiter_rad_value <- factor_rad_values[round(2*delimiter_rad_index)]

dataset$RAD.CAT <- cut(dataset$RAD, 
    breaks = c(-Inf, lower_delimiter_rad_value, upper_delimiter_rad_value, Inf), 
    labels = c('low', 'medium', 'high'),
    right = FALSE)

dataset$RM.CAT <- cut(dataset$RM, 
    breaks = c(-Inf, 4, 6, Inf), 
    labels = c('low', 'medium', 'high'),
    right = FALSE)

factor_indus_values <- sort(unique(dataset$INDUS))
delimiter_indus_index <- length(factor_indus_values)/4
lower_delimiter_indus_value <- factor_indus_values[round(delimiter_indus_index)]
mid_delimiter_indus_value <- factor_indus_values[round(2*delimiter_indus_index)]
upper_delimiter_indus_value <- factor_indus_values[round(3*delimiter_indus_index)]

dataset$INDUS.CAT <- cut(dataset$INDUS, 
    breaks = c(-Inf, lower_delimiter_indus_value, mid_delimiter_indus_value, upper_delimiter_indus_value, Inf), 
    labels = c('very low', 'low', 'high', 'very high'),
    right = FALSE)

dataset$LSTAT.CAT <- cut(dataset$LSTAT, 
    breaks = c(-Inf, 25, 50, 75, Inf), 
    labels = c('very low', 'low', 'high', 'very high'),
    right = FALSE)

dataset$ZN.CAT <- cut(dataset$ZN, 
    breaks = c(-Inf, 25, 50, 75, Inf), 
    labels = c('very low', 'low', 'high', 'very high'),
    right = FALSE)

str(dataset)

```



I'm going to try to find differences in shapes between the different levels of a predictor and the variable that we want to predict, the average price.



```{r}
pairs(data=dataset,~CRIM+ZN+INDUS+NOX+RM+AGE+DIS+RAD+TAX+PTRATIO+B+LSTAT+MEDV,pch=19,col=dataset$CHAS)
```

Within this plot we can see that there is a linear relationship between (NOX,DIS) and (LSTAT,MEDV)

Now, lets try to find groups through scatter plots.
```{r}
ggplot(data=dataset,aes(x=NOX,y=DIS))+geom_point(aes(colour=ZN.CAT))

ggplot(data=dataset,aes(y=MEDV,x=LSTAT))+geom_point(aes(colour=LSTAT.CAT))
```

Density curves

```{r}
#Given river
dataset %>% ggplot(aes(x=MEDV)) + geom_density(aes(color=CHAS))

#Given rooms
dataset %>% ggplot(aes(x=MEDV)) + geom_density(aes(color=RM.CAT,fill=RM.CAT))

dataset %>% ggplot(aes(x=MEDV)) + geom_density(aes(color=LSTAT.CAT))

dataset %>% ggplot(aes(x=NOX)) + geom_density(aes(color=ZN.CAT))
```

PCP plot

```{r}
color_rad_cat=c("blue","green","orange2")[dataset$RAD.CAT]
parcoord(dataset[,c(1,5,7:8,11,12,14)],col=color_rad_cat,var.label=TRUE)
```
We can see that for high RAD.CAT the variables more affected are CRIM, AGE and DIS

ANDREWS curve

```{r}
library(andrews)
par(mfrow=c(1,1))
andrews(df=as.data.frame(cbind(dataset[,c(1,5,7:8,11,12,14)],dataset$CHAS)),clr=8,ymax=4)
```



